Variance reduction for MCMC methods via martingale representations

### 2020-01-28 aap/Revision/vr-mcmc-revision-new.tex
constant step size

In this section we give a general discrete-time martingale representation for the Markov chain \eqref{eq:chain_gen} which is  used below to construct an efficient variance reduction algorithm. Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).  In particular, we have

@@

\begin{thm}\label{prop:29032018a1}
 Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$, with a Markov chain $(X_{p})_{p\geq 0}$ defined in \eqref{eq:chain_gen}.
Then,
for $p>j$, the following representation holds in \(L^2(P)\)
\label{eq:mart_repr}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X_{l-1})\phi_k\left(\xi_{l}\right),
\label{eq:coeff_mart}
a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.

@@

The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\cdots,p-1\}$

@@

@@

Next we show how  the representation \eqref{eq:mart_repr} can be used to efficiently reduce the variance of MCMC algorithms.  Let $(\gamma_{p})_{p\in\mathbb N}$ be a sequence of positive and non-increasing
step sizes
with $\sum_{p=1}^\infty \gamma_p=\infty$
and, for $n,l\in\mathbb{N}$, $n\le l$, we set
\Gamma_{n,l}=\sum_{p=n}^{l}\gamma_{p}.
 Consider a weighted average estimator $\pi_{n}^{N}(f)$ of the form

@@

\pi_{n}^{N}(f)=\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}f(X_{p}),\quad\omega_{p,n}^{N}=\gamma_{p+1}\Gamma_{N+2,N+n+1}^{-1},

@@

M_{K,n}^N(f) =\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\bar a_{l,k}(X_{l-1})\phi_k(\xi_{l}),

@@

\label{eq:first-expression-bar-a-l-k}
\bar{a}_{l,k}(x)
=\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,k}(x)=\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\phi_k(\xi_{l})\right|X_{l-1}=x\right].
Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\geq 1,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.

@@

should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).

@@

We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\label{eq:29032018a5}\\
&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and
\begin{align*}
\bar{a}_{l,\mathbf{k}}(x)
& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
\end{align*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
\\
\nonumber
&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
+\mathsf{Var}[\rho(f)].
\end{eqnarray}
Our analysis will be done under the following two assumptions.

@@

\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)	$ and any $y\in \mathbb{R}^d$ it holds 

@@

\langle D^2U(x)y,y\rangle \geq m_U\|y\|^2, \quad \|D^3 U(y)\| \leq M_U. 

@@

\mathrm{osc}(f)^2
\sum_{p=N+1}^{N+n}
\omega_{p,n}^N\,
\|Q_\gamma^{p}(x_0,\cdot)-\pi(\cdot)\|_{TV}^2,

@@

for $\gamma>0$, the Markov kernel $R_\gamma$

@@

R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,

@@

while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_\gamma^{k,l}$ is
$
Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
$
which, finally, provides the  measure
$Q_\gamma^{p}(x_0,\cdot)=Q_\gamma^{p,1}$
used in the right-hand side of~\eqref{eq:06042018a2}.

@@

on bounds in the total variation distance. For example, the following proposition holds.
Suppose that the assumption (H1) holds, then it holds
\|Q_\gamma^{p}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \sum_{k=r+1}^{p} \gamma_k^2+\exp\left(-\varkappa\sum_{k=r+1}^{p} \gamma_k\right),
for some $\varkappa>0,$ where $\lesssim$ stands for  inequality up to a constant depending on $L$ and $\mathcal G_N.$ In particular if $\gamma_k=k^{-\alpha}$ for some $\alpha\in (0.5,1], $ then 
\|Q_\gamma^{N+1,N+n}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \frac{1}{(N+1)^{2\alpha-1}}
provided that $N/n\to 0$ as $N,n\to \infty.$
\subsection{Conditional variance}

@@

is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
As for the estimator~\eqref{eq:29032018a3}, 
it follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
\label{cond_var_dec}
	For variance-reduced estimator~\eqref{eq:29032018a3}, defined for the Markov Chain~\eqref{eq:chain_gen} the following representation holds
\begin{equation}\label{eq:29032018a6}
\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]
%=\sum_{\min_i (k_i)\ge K+1}
=\sum_{\mathbf k\colon \|\mathbf k\|\ge K+1}
\sum_{l=N+1}^{N+n}
\mathsf{E}\left[\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}\right].
Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.

@@

Assume (H1) and (H2).
Suppose that $f$  is two times continuously differentiable

@@

|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad

@@

for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$ and $\sum_{k=1}^\infty \gamma^2_k<\infty.$ 
Then it holds under 

@@

\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},

@@

where \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)

@@

\begin{cor}

\section{Nonparametric regression}
The functions \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
We present a generic regression algorithm. Algorithm starts with estimating the functions $\bar{Q}_{l}$ for $l=N+1,\ldots,N+n,$ where
\begin{equation*}
\bar{Q}_{l}(x)=\sum_{p=l}^{N+n} \omega_{p,n}^{N} Q_{p,l}\left(x\right)=\omega_{l,n}^{N}f(x)+\mathsf{E}\left[\left.\sum_{p=l+1}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}=x\right] .
\end{equation*}
We first generate \(\NtrainPath\)  paths conditionally independent of the $\sigma$-algebra generated by the burn-in sequence $X_1,\dots,X_N$:
\[
\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
\]
of the chain \(X\)
(the so-called ``training paths'').
The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
\begin{equation}\label{eq:06042018a1}
{\widehat  Q}_{l}=\argmin_{\psi\in \Psi}\sum_{s=1}^{\NtrainPath} \left | \sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X^{(s)}_{p})-\psi(X^{(s)}_{l})\right|^2
\end{equation}
for \(l=N+1,\ldots, N+n-1.\)
where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
Next  we estimate the coefficients \( \bar a_{l,k}\) using
the formula
\begin{equation}
\label{eq:a_est_int}
\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi_l(x,\xi)\right)\right | \TrainSet\right] \,,
\end{equation}
where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\). Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories. In many cases integration can be done in closed analytical form.

 For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi_l(x,\xi)=x-\gamma_{l}\mu(x)+\sqrt{\gamma_{l}}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.

Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
\begin{eqnarray*}
\widehat M_{K,n}^N(f) =\sum_{1\leq  k \le K}\sum_{l=N+1}^{N+n}\widehat a_{l,k}(X_{l-1})\phi_k(\xi_{l}).
\end{eqnarray*}
Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in a sense that it
does not introduce any bias.

\section{Complexity analysis for ULA}
\label{sec:coeff}

The following result quantifies the error of estimating the functions \((\bar Q_l)\)
via the algorithm and its proof follows from Theorem~2.2 in \cite{audibert2011robust}.
\begin{thm}\label{berm:theorem:regression_cv}
 Suppose that for any \(l\in \{N+1,\ldots,N+n+1\},\)

@@

\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})-\bar Q_l(X_l)\right)^4\right]\leq \sigma_l^4,

@@

for some finite positive numbers \(\sigma_{N+1},\ldots, \sigma_{N+n}. \)
Furthermore, assume that \(\Psi=\mathrm{span}\{\psi_1,\ldots,\psi_D\},\) where the functions  \(\psi_1,\ldots,\psi_D\) are uniformly bounded and satisfy

@@

\max_{l}\sup_{g\in \Psi\setminus \{0\}} \|g\|^2_\infty/\mathsf{E}[g^2(X_l) | \mathcal{G}_N]\leq B<\infty.

@@

Then for any values of \(\varepsilon\) and \(\NtrainPath\) such that \(2/\NtrainPath\leq \varepsilon\leq 1\) and
\NtrainPath\gtrsim B^2\left[BD+\log(2/\varepsilon)+\frac{B^2D^2}{\NtrainPath}\right]
it holds with probability at least \(1-\varepsilon,\)
\begin{multline}
\mathsf{E}\left[ \left.\Bigl|\bar Q_l(X_l)
-\widehat Q_l(X_l)\Bigr|^2\right | \TrainSet \vee \mathcal{G}_N\right]
\lesssim  \sigma^2_l B
\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
 +
\inf_{\psi\in \Psi}
\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right], \label{eq:regr_error}
\end{multline}
for \(l=N+1,\ldots,N+n,
\) with \(\lesssim\) standing for inequality up to a universal multiplicative constant.
Now we are able to give  a bound for the difference between \( M_{K,n}^N\) and \(\widehat M_{K,n}^N.\)
\label{cor:dif_m}
Under  conditions of Theorem~\ref{berm:theorem:regression_cv}, we have with probability at least \(1-\varepsilon,\)
\label{eq:main_bound}
&\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet,\mathcal{G}_N\right] \\
&\phantom{\mathsf{E}\widehat{M}_{K,n}^{N}}  \lesssim K\left[\sum_{l=N+1}^{N+n} \sigma^2_l \right]
B
\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
\\ \nonumber
& \phantom{\mathsf{E}|\widehat{M}_{K,n}^{N}} + K \sum_{l=n+1}^{N+n}\inf_{\psi\in \Psi}
\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right].
Using the conditional Cauchy-Schwarz inequality and  orthonormality of \((\phi_k)_{k\geq 0}\), we derive
\mathsf{E}\left[\left.\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]\leq\mathsf{E}\left[\left.\left|\widehat{Q}_{l}\left(X_{l}\right)-\bar Q_{l}\left(X_{l}\right)\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]
By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
\begin{multline*}
\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]
\leq \sum_{1\leq k\le K}\sum_{l=N+1}^{N+n}\mathsf{E}\left[\left.\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right].
\end{multline*}
\subsection{Complexity analysis of variance reduction for the ULA algorithm}
In the case of ULA one can bound the quantities \(\sigma_l^2\) using  \(L^p\)-type Sobolev inequalities  (see \cite{Adamczak}), see Remark~\ref{rem:p-norm} in Section~\ref{sec:proofs}.
In particular, we can derive that  under conditions of Theorem~\ref{th:mr} with \(K=1,\)
\sigma_l^2\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},\quad l=N+1,\ldots,N+n+1,
where \(\lesssim \) stands for the inequality up to a multiplicative constant not depending on \(n\) and \(N.\) Using this inequality and combining \eqref{eq:main_bound} with Theorem~\ref{th:mr}, we get for the variance of \(\widehat \pi_{K,n}^N(f)=\pi_n^N(f)-\widehat M_{K,n}^N(f),\) with probability at least \(1-\varepsilon,\)
\mathsf{Var}\left[\left.\widehat \pi_{K,n}^N(f)\right | \TrainSet, \mathcal{G}_N\right]&\lesssim & \frac{n KB}{\Gamma^2_{N+2,N+n+1}}
\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
&& + \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
\label{eq:ula_red_var}
&& +
K \sum_{l=n+1}^{N+n}\inf_{\psi\in \Psi}
\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N \right].
In order to assess the complexity of the proposed algorithm,  we first prove that under some conditions the coefficients \(a_{p,l,\mathbf{k}}\) decay exponentially fast as \(|p-l|\to \infty.\)
\begin{lem}
\label{a_decay}
Suppose that \(f,\mu\in C^1(\mathbb{R}^d),\) \(0<b_\mu I\leq\jac{\mu}(x)\leq B_\mu I\) and \(B_\mu\gamma_{l}\leq 1,\) then
\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\, \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right),\quad \mathbf{k}\in \mathbb{N}^d\setminus \{\mathbf{0}\}.
\end{lem}
\begin{cor}
Assume that
\label{eq:gamk-decay}
\gamma_k=\gamma_1 k^{-\alpha},\quad \alpha\in (0,1),
 then
\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_1} l^{-\alpha/2}\, \|\nabla f\|_\infty\exp\left(-c b_{\mu}\gamma_1 (p^{1-\alpha}-l^{1-\alpha}) \right)
for some constant \(c>0.\)
Suppose now that the chain is close to stationarity, then \(Q_{p,l}\) are functions of \(p-l\) only.
Furthermore,  according to Lemma~\ref{a_decay}, \(a_{p,l,\mathbf{k}}\) are exponentially small for \(p-l\) large.
As a result we only need to compute a logarithmic (in \(n\)) number of functions \(Q_{p,l}.\)
Hence the cost of computing the estimates  \(\widehat {a}_{p,l,\mathbf{k}}(x)\) for \(l=N+1,\ldots,N+n,\) and \(\|\mathbf{k}\|\leq K\) using regression on \(\Psi,\) is of order

@@

+\gamma_{p}

>>>>

In this section we give a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used below to construct an efficient variance reduction algorithm. Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).  In particular, we have

@@

\begin{thm}\label{prop:29032018a1}
 Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some kernels $(\Phi_l)_{l\geq 1}.$  
Then,
for $p>j$, the following representation holds in \(L^2(P)\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where
\begin{eqnarray}
\label{eq:coeff_mart}
a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
\end{eqnarray}
\end{thm}
\begin{cor}
If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form 
\begin{eqnarray*}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
with 
\begin{eqnarray*}
\bar a_{r,k}(x)=\mathsf{E}\left[\left.f(X_{r})\phi_k\left(\xi_{1}\right)\right|X_{0}=x\right], \quad r, k\in \mathbb{N}.
\end{eqnarray*}

@@

The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$

@@

In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have 
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
with  \(Q_{r}(x)=\mathsf{E}\left[\left.f(X_{r})\right|X_{0}=x\right],\) $r\in \mathbb{N}.$

@@

Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form

@@

\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X_{p}),

@@

M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right),

@@

A_{s,k}(x)=\frac{1}{s}\sum_{r=0}^{s}\bar a_{r,k}(x).
Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq l,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.

@@

should be small for \(K\) large enough. Indeed, due to orthonormality  we have
\mathsf{Var}[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\mathsf{E}[A^2_{n-l,k}(X_{N+l-1})]
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).

@@

In this section we perform the convergence analysis of the ULA algorithm. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma=\gamma_1=\gamma_2=\ldots.$
We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
with \(\|\mathbf{k}\|=\max_{i} k_i.\) 
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
+\mathsf{Var}[\rho(f)].
Our analysis will be done under the following two assumptions.

@@

\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds 

@@

\langle D^2U(x),x\rangle \geq \frac{m_U}{2}\|x\|^2. 

@@

\frac{\mathrm{osc}(f)^2}{n}\sum_{p=N+1}^{N+n}
\|\delta_x Q_\gamma^{p}-\pi(\cdot)\|_{TV}^2,

@@

for $\gamma>0$, the Markov kernel $Q_\gamma$

@@

\mathcal{Q}_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy.

@@

@@

on bounds in the total variation distance. It is known  that under assumptions {\bf (H1)} and {\bf (H2)}  the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. Moreover, as shown in Proposition (Appendix) \begin{eqnarray}
\label{eq:bias}
\|\delta_x\mathcal{Q}_{\gamma}^{n}-\pi_\gamma\|_{TV} \leq C_1\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right), \quad \|\pi-\pi_\gamma\|_{TV} \leq C_2 \sqrt{\gamma}
for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and  constants $C_1,$ $C_2>0$ independent of $\gamma$  and $n.$
\subsection{Variance}

@@

under {\bf (H1)} and {\bf (H2)} is derived in Section.  In particular, for any bounded function $f,$
\label{eq:var-mc}
\mathsf{Var}[\pi_n^N(f)]\lesssim \frac{1}{n\gamma},
where $\lesssim$ stands for inequality up to a constant depending on constants $L_U,$ $m_U,$  $M_U$ and $\|f\|_\infty.$ 
One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$ 

@@

Assume {\bf (H1)} and {\bf (H2)}.
Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K\geq 2$ times continuously differentiable

@@

\label{eq:smooth-mu}
|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} \mu (x)|\leq B_\mu, \quad 

@@

for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  K.\)  Then it holds 

@@

\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{\gamma^{K-2}}{n},

@@

where \(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)

@@

Let us sketch the main steps of the proof. First using integration by parts we prove that 
\label{eq:Ask-part}
A_{s,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_s) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]
where $F_s(X_0,Z_1,\ldots,Z_s)=\frac{1}{s}\sum_{r=0}^s f(X_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{s,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma}{2}\right)^{|I|K}
\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
\right),
where the sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$ and 

@@

\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).

@@

 Finally we show via  the Poincare inequality that under smoothness assumption \eqref{eq:smooth-mu} it holds
\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
\right)\leq W_s(x)
for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
\section{Proofs}
\label{sec:proofs}
\subsection{Proof of Theorem~\ref{th:mr}}
We start with introducing some notations.
For $m\in\mathbb N$, a smooth function
$h\colon\mathbb R^{d\times m}\to\mathbb R$
with arguments being denoted
$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
and $j\in\{1,\ldots,m\}$,
we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
with respect to the components of~$y_j$:
\partial^{\mathbf k}_{j} h(y_1,\ldots,y_m)
:=\partial^{k_d}_{y_j^d}
\ldots
\partial^{k_1}_{y_j^1}
h(y_1,\ldots,z_m),
\quad y_j=(y_j^1,\ldots,y_j^d).
In the particular case $m=1$ we can drop
the subscript $y_1$ in that notation.
For $l\le p$, we have the representation

@@

X_{p}=G_{p}(X_0,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}),

@@

where the function \(G_{p}:\) \(\mathbb{R}^{d\times(p+1)}\to \mathbb{R}^{d}\) is defined as
\label{eq:definition-G-p-l}
G_{p}(x,y_1,\ldots,y_p):=\Phi(\cdot,y_{p})\circ\Phi(\cdot,y_{p-1})\circ\ldots\circ\Phi(x,y_{1})

@@

+\gamma
